A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2019-07-01 paper/paper.Rnw
Fix typos

Many convolutional neural networks have been trained on a subset of images from an online image database called ImageNet \citep{dengImageNetLargeScaleHierarchical}. In particular, there is a standard set of images from ImageNet, consisting of 1.2 million images that span 1,000 catetories, that is commonly used in computer vision tasks. CNNs trained on ImageNet are optimized for general human-like vision, that is, the ability to recognize a large set of different features simultaneously. As a result, the base of networks trained on these sets are often used for transfer learning, because the initial layers are broadly generalizable to a wide variety of more specific image labeling tasks. In the next section, we will introduce the pre-trained network VGG16, pictured in \autoref{fig:VGG16-structure}, and examine a selection of its filters (shown in \autoref{fig:vgg16layers}) that demonstrate the levels of detail in the initial layers of a model trained on generic data.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{images/vgg16-shoe-nolabel}
\caption[VGG16 structure]{VGG16 consists of five convolutional blocks that make up the model base. Each convolutional block contains an increasing number of increasingly complex \mt{feature-detecting filters}. After the convolutional blocks, the fully connected layers of the model head are used to make global connections between separate features.}\label{fig:VGG16-structure}
\end{figure}

@@

Although the filters of VGG16 have been trained and optimized on the ImageNet data, the large number of filters span a wide variety of shapes and features, many of which are useful when examining tread patterns. Just as humans can detect the features of an unfamiliar object (consituent parts, edges, textures, and corners), a CNN trained on a wide set of objects can be used to classify unrelated images, such as shoe soles.

@@

\paragraph{Overall Accuracy}
For each image in the test set, the ``true" labels are nine human-assigned labels of 0 or 1 corresponding to the presence or absence of each of the nine shape categories in the image. When the CNN predicts which shapes are in the image, however, it assigns a probability between 0 and 1 for each shape category, with a total of nine probabilities per test image. Thus, to determine the accuracy of the model predictions, we must select a threshold for each label to discretize the model predictions. If the threshold is too low, moderate probabilites will be more likely to exceed the threshold, which increases the true positive and false positive predictions in the test data. Conversely, if the threshold is too high, moderate probabilities will more often fail to meet the threshold, which increases the number of true and false negative predictions. An appropriate threshold must be chosen to produce a high true positive rate while also ideally keeping the false positive rate low. One way to set an optimal threshold is to utilize Receiver Operating Characteristic (ROC) curves, a type of diagnostic plot which compare the false positive rate to the true positive rate for a classification method. \autoref{fig:overall-roc} shows the ROC curve for our model across all classes, and \autoref{fig:class-roc} shows the curve for each class. ROC curves can be summerized by the Area Under the Curve (AUC), which quantifies the overall accuracy of the model. Perfect prediction would be indicated by a right angle along the upper left corner of the plot, with a corresponding AUC of 1, and a diagonal line with an AUC of 0.5 would indicate that the classification method performs no better than random chance.

@@

\paragraph{Class Accuracy}
The ROC curves provide only a general summary of model accuracy; to learn more about how specific images are classified, we can view images from the test set along with the nine class predictions assigned by the model. \autoref{fig:exemplars} and \autoref{fig:circle-text-consistency} show model predictions for a selection of images. Each image shown in the following examples belongs to either the validation or test data, so none of the images were used during model training. In these figures, the values in each row are the predicted probabilities for the nine classes in the corresponding image, such that each column represents predictions for a given class. The saturation of blue color corresponds to the strength of the predicted probability, and the navy blue border around an image indicates that the predicted probability is above the EER threshold for the given class. Note that the EER is not the same for all classes; in other words, the same probablity value assigned across different categories may yield different conclusions about whether the model predicts the shape is present or absent in the image.

The first rows of \autoref{fig:exemplars} are images which contain only one of the nine classes. Most of the predictions for these images are promising; there are some false-positive labels that indicate confusions between circle/text and polygon/quad, but all of the true shapes are corectly identified by the model above their respective EER. The next image belongs to the ``other" category, as it does not contain any of the nine classes; the model correctly assigns near-zero probabilities to each class. The final rows are of images that belong to more than one class category. Again, predictions are largely accurate, with most of the prominent shapes correctly identified above EER. The most notable misclassifications in these images are the false positive predictions of triangle in the image with green text and lines in the Adidas logo. Although the Adidas logo is technically made up of adjacent quadrilaterals, the prediction of a triangle is reasonable when considering the proximity of those quadrilaterals and the shape's features as a whole, and the relatively low predicted probability suggests that the model recognizes that this shape is ambiguous. All other predictions for these images are accurate given our operational definition of class categories.

@@

While figure \ref{fig:exemplars} shows representative examples of classifications across multiple shape categories, it is also useful to examine many cases within a given shape to get a sense of model consistency. \autoref{fig:circle-text-consistency} shows a number of images containing circles and/or text. The top four rows show text contained within a separate, distinct circle shape, and the next five rows show circles represented as the letter ``o" in text. The remaining images contain only one of either text or circles, as well as other possible shape categories. In particular, the first two of these images contain text that use the capital letter ``G", the next two images contain no circular shapes, and the final 3 images contain pure circles that do not appear to belong to text. The model is able to identify the circle in all cases where a circle is truly present, regardless of whether the circle is part of the text or it is distinct; however, the circle prediction is also strong in three of the four images that do not contain a circle, which indicates that the model is having dificulty distinguishing the two categories. While the circle predictions are often above the EER for letters such as G and S, these predictions are generally much lower than those where distinct circles or Os are seen, indicating that there are some indications that features are not as strong in the quasi-circle shapes, but that the model is not able to completely separate Os from C, S, and G shapes which are also very circular. This confusion is complicated by the frequency of images which contain both circles and text: text is present in \Sexpr{100 * round(sum(as.logical(grepl("circle", dfunion$name) & grepl("text", dfunion$name))) / sum(grepl("circle", dfunion$name)), 2)}\% of our labeled images containing circles, and circles are present in \Sexpr{100 * round(sum(as.logical(grepl("circle", dfunion$name) & grepl("text", dfunion$name))) / sum(grepl("text", dfunion$name)), 2)}\% of our images containing text. In addition, the distinction between a circle and an ``o" comes from the context around the shape, which is much more difficult for the model to parse than simply detecting a shape. More importantly, this behavior does not suggest poor model performance as much as a weakness of the classification scheme, which will be discussed in a later section.

@@

The horizontal band in \autoref{fig:confusion-matrix} indicates that quadrilaterals are predicted more often than they should be for every true label. Similarly, polygons and triangles produce a moderate number of false positive predictions in other categories, as evidenced by the vertical bands for these categories. It is also apparent that circles and text are commonly confused, which supports our earlier obsesrvartions from \autoref{fig:circle-text-consistency} that these two classes are difficult for the model to distinguish, in part because circles are relatively common components of latin script characters.

@@

\paragraph{Image Contrast}
As described earlier, the images used for this research were obtained from online retailers before being labeled and cropped. One consequence of using outsole images instead of impressions is the impact of colors in the image. Features from outsoles that are primarily black or white in color are typically not as prominent as features from images with more varied color schemes. As a result, the features in low contrast images are not easily resolved by the filters; this is propogated through both model training and prediction of new images. In other words, the model `learns' that when there are few meaningful features with which to determine which shapes are present, the best strategy for minimizing loss is to predict diffuse probabilities across many classes, as seen in \autoref{fig:low-contrast-consistency}.

@@

\mt{Figures \ref{fig:ugg-heatmap}-\ref{fig:dc-heatmap} show a number of images alongside heatmaps for the four classes with the highest output probabilities for each image. \autoref{fig:ugg-heatmap} shows predictions and heatmaps for the logo from an UGG shoe. The model correctly identifies triangles and text, and references appropriate regions of the image. The prediction of a circle is shown to be associated with the letter 'G'; although this prediction is technically incorrect, it indicates that the model has learned to associate curved shapes with circles. This association is again seen in the high predicted probability of circle in \autoref{fig:seychelles-heatmap}. In this case, the prediction of text is again correct,, but the prediction of circle is not---there are very round shapes, but they do not form a closed figure that could be identified as a circle. \autoref{fig:dc-heatmap} shows a DC shoe with repeating instances of a shape that is a combination of a circle and quadrilateral, but does not technically fall into either of those categories by our definition. The model assigns high probabilities of circle and quadrilateral and indicates that the proper image features are being used for this classification. Additionally, the prediction of polygon is incorrect but reasonable considering that many small hexagons are indistinguishable from circles. These heatmaps can confirm that a number of misclassifications are a result of overlapping features between classes, and that the model is using reasonable features to make predictions.}

@@

<<current-heatmaps, eval = T, out.width = ".95\\textwidth", fig.width = 8, fig.height = 6, fig.cap = c("A section of an UGG shoe containing the logo. The model correctly identifies triangles and text (and references the correct portion of the image in both cases). The model identifies a circle as well; while it could be argued that the triangles form an implicit circle, the heatmap demonstrates that the circular portion of the image is the 'G', which would not be a closed figure labeled as a circle. The model's prediction of the presence of a circle is understandable, but incorrect.\\label{fig:ugg-heatmap}", "In this image of a Seychelles shoe, the model identifies text and circle. The prediction of text is correct; the prediction of circle is not---there are very round shapes, but they do not form a closed figure that could be identified as a circle.\\label{fig:seychelles-heatmap}", "This image from a DC shoe contains repeating instances of a shape that appears to be a combination of a circle and quadrilateral, but does not technically fall into any of the nine categories used in our labeling scheme. The model assigns high probabilities of circle and quadrilateral and indicates that the proper image features are being used for this classification. The prediction of polygon is incorrect but reasonable considering that many small hexagons are indistinguishable from circles.\\label{fig:dc-heatmap}"), echo = F, fig.pos = "!p">>=

>>>>

Many convolutional neural networks have been trained on a subset of images from an online image database called ImageNet \citep{dengImageNetLargeScaleHierarchical}. In particular, there is a standard set of images from ImageNet, consisting of 1.2 million images that span 1,000 categories, that is commonly used in computer vision tasks. CNNs trained on ImageNet are optimized for general human-like vision, that is, the ability to recognize a large set of different features simultaneously. As a result, the base of networks trained on these sets are often used for transfer learning, because the initial layers are broadly generalizable to a wide variety of more specific image labeling tasks. In the next section, we will introduce the pre-trained network VGG16, pictured in \autoref{fig:VGG16-structure}, and examine a selection of its filters (shown in \autoref{fig:vgg16layers}) that demonstrate the levels of detail in the initial layers of a model trained on generic data.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{images/vgg16-shoe-nolabel}
\caption[VGG16 structure]{VGG16 consists of five convolutional blocks that make up the model base. Each convolutional block contains an increasing number of increasingly complex feature-detecting filters. After the convolutional blocks, the fully connected layers of the model head are used to make global connections between separate features.}\label{fig:VGG16-structure}
\end{figure}

@@

Although the filters of VGG16 have been trained and optimized on the ImageNet data, the large number of filters span a wide variety of shapes and features, many of which are useful when examining tread patterns. Just as humans can detect the features of an unfamiliar object (constituent parts, edges, textures, and corners), a CNN trained on a wide set of objects can be used to classify unrelated images, such as shoe soles.

@@

\paragraph{Overall Accuracy}
For each image in the test set, the ``true" labels are nine human-assigned labels of 0 or 1 corresponding to the presence or absence of each of the nine shape categories in the image. When the CNN predicts which shapes are in the image, however, it assigns a probability between 0 and 1 for each shape category, with a total of nine probabilities per test image. Thus, to determine the accuracy of the model predictions, we must select a threshold for each label to discretize the model predictions. If the threshold is too low, moderate probabilities will be more likely to exceed the threshold, which increases the true positive and false positive predictions in the test data. Conversely, if the threshold is too high, moderate probabilities will more often fail to meet the threshold, which increases the number of true and false negative predictions. An appropriate threshold must be chosen to produce a high true positive rate while also ideally keeping the false positive rate low. One way to set an optimal threshold is to utilize Receiver Operating Characteristic (ROC) curves, a type of diagnostic plot which compare the false positive rate to the true positive rate for a classification method. \autoref{fig:overall-roc} shows the ROC curve for our model across all classes, and \autoref{fig:class-roc} shows the curve for each class. ROC curves can be summarized by the Area Under the Curve (AUC), which quantifies the overall accuracy of the model. Perfect prediction would be indicated by a right angle along the upper left corner of the plot, with a corresponding AUC of 1, and a diagonal line with an AUC of 0.5 would indicate that the classification method performs no better than random chance.

@@

\paragraph{Class Accuracy}
The ROC curves provide only a general summary of model accuracy; to learn more about how specific images are classified, we can view images from the test set along with the nine class predictions assigned by the model. \autoref{fig:exemplars} and \autoref{fig:circle-text-consistency} show model predictions for a selection of images. Each image shown in the following examples belongs to either the validation or test data, so none of the images were used during model training. In these figures, the values in each row are the predicted probabilities for the nine classes in the corresponding image, such that each column represents predictions for a given class. The saturation of blue color corresponds to the strength of the predicted probability, and the navy blue border around an image indicates that the predicted probability is above the EER threshold for the given class. Note that the EER is not the same for all classes; in other words, the same probability value assigned across different categories may yield different conclusions about whether the model predicts the shape is present or absent in the image.

The first rows of \autoref{fig:exemplars} are images which contain only one of the nine classes. Most of the predictions for these images are promising; there are some false-positive labels that indicate confusions between circle/text and polygon/quad, but all of the true shapes are correctly identified by the model above their respective EER. The next image belongs to the ``other" category, as it does not contain any of the nine classes; the model correctly assigns near-zero probabilities to each class. The final rows are of images that belong to more than one class category. Again, predictions are largely accurate, with most of the prominent shapes correctly identified above EER. The most notable misclassifications in these images are the false positive predictions of triangle in the image with green text and lines in the Adidas logo. Although the Adidas logo is technically made up of adjacent quadrilaterals, the prediction of a triangle is reasonable when considering the proximity of those quadrilaterals and the shape's features as a whole, and the relatively low predicted probability suggests that the model recognizes that this shape is ambiguous. All other predictions for these images are accurate given our operational definition of class categories.

@@

While figure \ref{fig:exemplars} shows representative examples of classifications across multiple shape categories, it is also useful to examine many cases within a given shape to get a sense of model consistency. \autoref{fig:circle-text-consistency} shows a number of images containing circles and/or text. The top four rows show text contained within a separate, distinct circle shape, and the next five rows show circles represented as the letter ``o" in text. The remaining images contain only one of either text or circles, as well as other possible shape categories. In particular, the first two of these images contain text that use the capital letter ``G", the next two images contain no circular shapes, and the final 3 images contain pure circles that do not appear to belong to text. The model is able to identify the circle in all cases where a circle is truly present, regardless of whether the circle is part of the text or it is distinct; however, the circle prediction is also strong in three of the four images that do not contain a circle, which indicates that the model is having difficulty distinguishing the two categories. While the circle predictions are often above the EER for letters such as G and S, these predictions are generally much lower than those where distinct circles or Os are seen, indicating that there are some indications that features are not as strong in the quasi-circle shapes, but that the model is not able to completely separate Os from C, S, and G shapes which are also very circular. This confusion is complicated by the frequency of images which contain both circles and text: text is present in \Sexpr{100 * round(sum(as.logical(grepl("circle", dfunion$name) & grepl("text", dfunion$name))) / sum(grepl("circle", dfunion$name)), 2)}\% of our labeled images containing circles, and circles are present in \Sexpr{100 * round(sum(as.logical(grepl("circle", dfunion$name) & grepl("text", dfunion$name))) / sum(grepl("text", dfunion$name)), 2)}\% of our images containing text. In addition, the distinction between a circle and an ``o" comes from the context around the shape, which is much more difficult for the model to parse than simply detecting a shape. More importantly, this behavior does not suggest poor model performance as much as a weakness of the classification scheme, which will be discussed in a later section.

@@

The horizontal band in \autoref{fig:confusion-matrix} indicates that quadrilaterals are predicted more often than they should be for every true label. Similarly, polygons and triangles produce a moderate number of false positive predictions in other categories, as evidenced by the vertical bands for these categories. It is also apparent that circles and text are commonly confused, which supports our earlier observations from \autoref{fig:circle-text-consistency} that these two classes are difficult for the model to distinguish, in part because circles are relatively common components of Latin script characters.

@@

\paragraph{Image Contrast}
As described earlier, the images used for this research were obtained from online retailers before being labeled and cropped. One consequence of using outsole images instead of impressions is the impact of colors in the image. Features from outsoles that are primarily black or white in color are typically not as prominent as features from images with more varied color schemes. As a result, the features in low contrast images are not easily resolved by the filters; this is propagated through both model training and prediction of new images. In other words, the model `learns' that when there are few meaningful features with which to determine which shapes are present, the best strategy for minimizing loss is to predict diffuse probabilities across many classes, as seen in \autoref{fig:low-contrast-consistency}.

@@

Figures \ref{fig:ugg-heatmap}-\ref{fig:dc-heatmap} show a number of images alongside heatmaps for the four classes with the highest output probabilities for each image. \autoref{fig:ugg-heatmap} shows predictions and heatmaps for the logo from an UGG shoe. The model correctly identifies triangles and text, and references appropriate regions of the image. The prediction of a circle is shown to be associated with the letter ``G"; although this prediction is technically incorrect, it indicates that the model has learned to associate curved shapes with circles. This association is again seen in the high predicted probability of circle in \autoref{fig:seychelles-heatmap}. In this case, the prediction of text is again correct,, but the prediction of circle is not---there are very round shapes, but they do not form a closed figure that could be identified as a circle. \autoref{fig:dc-heatmap} shows a DC shoe with repeating instances of a shape that is a combination of a circle and quadrilateral, but does not technically fall into either of those categories by our definition. The model assigns high probabilities of circle and quadrilateral and indicates that the proper image features are being used for this classification. Additionally, the prediction of polygon is incorrect but reasonable considering that many small hexagons are indistinguishable from circles. These heatmaps can confirm that a number of misclassifications are a result of overlapping features between classes, and that the model is using reasonable features to make predictions.

@@

<<current-heatmaps, eval = T, out.width = ".95\\textwidth", fig.width = 8, fig.height = 6, fig.cap = c("A section of an UGG shoe containing the logo. The model correctly identifies triangles and text (and references the correct portion of the image in both cases). The model identifies a circle as well; while it could be argued that the triangles form an implicit circle, the heatmap demonstrates that the circular portion of the image is the ``G\", which would not be a closed figure labeled as a circle. The model's prediction of the presence of a circle is understandable, but incorrect.\\label{fig:ugg-heatmap}", "In this image of a Seychelles shoe, the model identifies text and circle. The prediction of text is correct; the prediction of circle is not---there are very round shapes, but they do not form a closed figure that could be identified as a circle.\\label{fig:seychelles-heatmap}", "This image from a DC shoe contains repeating instances of a shape that appears to be a combination of a circle and quadrilateral, but does not technically fall into any of the nine categories used in our labeling scheme. The model assigns high probabilities of circle and quadrilateral and indicates that the proper image features are being used for this classification. The prediction of polygon is incorrect but reasonable considering that many small hexagons are indistinguishable from circles.\\label{fig:dc-heatmap}"), echo = F, fig.pos = "!p">>=
